Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-19 manuscript/manuscript.tex
removed old comment
added citations to model feature section

Here we consider how purifying selection is handled by codon substitution models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use\jdbcomment{find a citation for this and replace word ``accurate'' with something else if appropriate. Perhaps this one or something cited within \url{https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4620419/}}.
Codon substitution models distinguish between two types of substitutions: synonymous and nonsynonymous.
The relative rate of these substitutions is referred to as dN/dS or $\omega$.
In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called M0.
We will refer to the M0 variant of the \citet{goldman1994codon} model simply as GY94.
The gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.

@@

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94 M5 model with the typical F3X4 stationary state\jdbcomment{add citation, maybe just GY94 paper?} has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9500 parameters for a 500-residue protein.
It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so they do not need to be fit from phylogenetic data.
Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

@@

\begin{table}[t!]
\caption{\label{tab:empirical_data}
Performance comparison for the models in~\ref{fig:empirical_trees}. 
All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}
The mean $\omega$ value from each $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation, otherwise simply the fit, gene-wide $\omega$. 
Each ExpCM has a stringency parameter $>1$.
\skhcomment{GY94+$\Gamma\omega$ has one $\omega$ of 0 because of rounding.}
\skhcomment{check LL values}
} 
     \begin{tabular}{cccccccccc}
        \hline
         Model & {\shortstack{Site-specific\\ stationary state}} & $\Gamma\omega$ & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{stringency\\ parameter}}\\ \hline
       	ExpCM + $\Gamma\omega$ (H1+H3 avg) & + & + & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
	ExpCM (H1+H3 avg) & + & - &  950 & -49227 & 0.15 & 1.78\\
	ExpCM + $\Gamma\omega$ (H1) & + & + & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
	ExpCM + $\Gamma\omega$ (H3) & + & + & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
	ExpCM (H1) & + & - & 2556 & -50030 &  0.13 & 1.22\\
	ExpCM (H3) & + & - &  3197 & -50350 & 0.12 & 1.45\\
	GY94 + $\Gamma\omega$ & - & + & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & - \\
	GY94 & - & - & 7625 & -52560  & 0.07 & -\\
      \end{tabular}
\end{table}

>>>>

Here we consider how purifying selection is handled by codon substitution models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
Codon substitution models distinguish between two types of substitutions: synonymous and nonsynonymous.
The relative rate of these substitutions is referred to as dN/dS or $\omega$.
In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called M0.
We will refer to the M0 variant of the \citet{goldman1994codon} model simply as GY94.
The gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.

@@

However, the increased realism of mutation-selection models comes at the cost of an increased number of parameters. 
Codon substitution models with uniform stationary states have only a modest number of parameters that must be fit from the phylogenetic data.
For instance, a GY94 M5 model with the typical F3X4 stationary state~\citep{yang1994maximum} has 12 parameters: two describing the shape of the gamma distribution over $\omega$, a transition-transversion rate, and 9 parameters describing nucleotide composition of the stationary state.
However, mutation-selection models must additionally specify 19 parameters defining the stationary state for \emph{each} site (there are 20 amino acids whose frequencies are constrained to sum to one).
This corresponds to $19\times L$ parameters for a protein of length $L$, or 9500 parameters for a 500-residue protein.
It is challenging to obtain values for these parameters without overfitting the data~\citep{rodrigue2013statistical}.
Here we will primarily use experimentally informed codon models (ExpCM)~\citep{bloom2014experimentally, hilton2017phydms, bloom2017identification} which define values for these parameters \textit{a priori} from deep mutational scanning experiments~\citep{fowler2014deep} so they do not need to be fit from phylogenetic data.
Therefore, the number of remaining free parameters for an ExpCM are similar to a non-site-specific substitution model.
Alternative strategies of obtaining parameters for site-specific stationary states via Bayesian\skhcomment{cite} or maximum-likelihood estimation\skhcomment{cite} are discussed in the last section of the Results.

@@

\begin{table}[t!]
\caption{\label{tab:empirical_data}
Performance comparison for the models in~\ref{fig:empirical_trees}. 
All ExpCMs describe the evolution of HA better than the GY94 models, as evaluated by the Akaike information criteria~\citep[$\Delta$AIC,][]{posada2004model}
The mean $\omega$ value from each $K=4$ bins is shown for the models with $\Gamma\omega$ rate variation, otherwise simply the fit, gene-wide $\omega$. 
Each ExpCM has a stringency parameter $>1$.
\skhcomment{GY94+$\Gamma\omega$ has one $\omega$ of 0 because of rounding.}
} 
     \begin{tabular}{cccccccccc}
        \hline
         Model & {\shortstack{Site-specific\\ stationary state}} & $\Gamma\omega$ & $\Delta$AIC & {\shortstack{Log\\ Likelihood}} & {\shortstack{$\omega$\\ (implied dN/dS)}} & {\shortstack{stringency\\ parameter}}\\ \hline
       	ExpCM + $\Gamma\omega$ (H1+H3 avg) & + & + & 0 & -48751 & 0.19,  0.50,  0.90,  1.86 &  1.70\\
	ExpCM (H1+H3 avg) & + & - &  950 & -49227 & 0.15 & 1.78\\
	ExpCM + $\Gamma\omega$ (H1) & + & + & 1306 & -49404  & 0.13 ,  0.44,  0.91,  2.16 & 1.12\\
	ExpCM + $\Gamma\omega$ (H3) & + & + & 1737 & -49620 & 0.09,  0.33,  0.72,  1.77 & 1.28\\
	ExpCM (H1) & + & - & 2556 & -50030 &  0.13 & 1.22\\
	ExpCM (H3) & + & - &  3197 & -50350 & 0.12 & 1.45\\
	GY94 + $\Gamma\omega$ & - & + & 4719 & -51106 & 0.00,  0.03,  0.08,  0.26 & - \\
	GY94 & - & - & 7625 & -52560  & 0.07 & -\\
      \end{tabular}
\end{table}
